---
title: "Portfolio 1 - Experimental Methods 1 2019"
author: "Søren Orm Hansen"
date: "9/22/2019"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE)

library(pacman)

pacman::p_load(tidyverse, data.table, lme4, car, ggplot2, pastecs, psych, plotrix)

```


## Portfolio exam - Part 1

#### Write a short data mining report on the CogSci Intro Week Personality Test Data in which you answer the following questions in prose, code and graphs:

First of all, you have to import the personality data. If the dataset is in your working directory, then you can just load it into R. Otherwise you will have to change your working directory.

```{r load_data}
df <- read.table("PTD.txt", header = TRUE, sep = ",")

```

Once you are done loading the data, you can start working on the questions below.

#### Question 1

Who can hold their breath longer on average, those with right or left ocular dominance? Plot the data using ggplot2 to find out. The plots should include error bars (depicting the standard error of the means). Then use the mean() and sd() functions to find mean and standard deviation of the two eye dominance groups.


```{r question1}
#Since we are only interested in people with either right or left ocular dominance, I will start by filtering the data, so I only have people with right or left ocular dominance.

filter_ocular_dom <- filter(df, ocular_dom != "Both")

#barplot with SEM-errorbars - average breath hold by ocular dominance

ggplot(filter_ocular_dom,
       aes(x = ocular_dom, y = breath_hold, fill = ocular_dom)) +
  geom_bar(
    stat = 'summary',
    fun.y = mean,
    width = 0.5,
    color = "Black"
  ) +
  stat_summary(
    fun.data = mean_cl_normal,
    geom = "errorbar",
    color = "Black",
    width = 0.1
  ) +
  theme_minimal() + 
  scale_fill_brewer(palette = "Blues") +  
  labs(title = "Barplot of average breath hold by ocular dominance", 
       x = "Ocular Dominance", 
       y = "Breath Hold"
  )

#Finding the mean, standard deviation, the 95% confidence interval, and the numbers of observations in each category

summary_breath_byod <- filter_ocular_dom %>%
  group_by(ocular_dom) %>%
  summarise(
    mean(breath_hold),
    sd(breath_hold),
    SEM_max = mean(breath_hold) + std.error(breath_hold)*1.96,
    SEM_min = mean(breath_hold) - std.error(breath_hold)*1.96,
    Count = n()
  )
summary_breath_byod
```

Explain your results in plain terms here: 

Our data tells us, that people with right ocular dominance in our sample can hold their breath longer on average (57.7 seconds) than people in our sample with left ocular dominance (53.4 seconds). 

To know how certain we are of the two averages we've found, we look at the 95% confidence or the interval in which we are 95% certain that the true average of the population is. In the table I have provided the maximum (SEM_max) and minimum (SEM_min) values for the 95% confidence interval for both right and left ocular dominance. We see that these two intervals overlap. This indicates, that the difference between the two averages is not statistically significant - that the difference is likely to just have occured by chance.

The two bars in the graph indicate the means of the two groups and the 'capital I's' on top of the bars indicate the 95% confidence intervals. Here we see the difference in the two means and the overlapping 95% confidence intervals.

Thus, we cannot conclude from the data in our sample whether people with right eye or left eye dominance can hold their breath longer


#The standard diviation was found to be 21.29 for people with left ocular dominance and 20.63 for people with right ocular dominance. The standard diviation tells us something about the variance, and in this instance it is quite large.

#Do we use the z or t test when the sample size is 21?

---

#### Question 2

Who likes silence vs. noise best? Males or females? Also in this case you should plot the data using ggplot2 (including error bars), then use the mean() and sd() functions to find mean and standard deviation of the two genders.

```{r question2}
#Boxplot with mean and errorbars - prefered sound level by gender

ggplot(df,
       aes(x = gender, y = sound_level_pref, fill = gender)) +
  geom_bar(
    stat = 'summary',
    fun.y = mean,
    width = 0.5,
    color = "Black"
  ) +
  stat_summary(
    fun.data = mean_cl_normal,
    geom = "errorbar",
    color = "Black",
    width = 0.1
  ) +
  theme_minimal() + 
  scale_fill_brewer(palette = "Blues") +  
  labs(title = "Boxplot of prefered sound level by gender", 
       y = "Prefered Sound Level (0-100)", 
       x = "Gender"
  )

#Finding the mean, standard deviation, the 95% confidence interval, and the numbers of observations in each category

summary_soundlvl_bygender <- df %>%
  group_by(gender) %>%
  summarise(
    Mean = mean(sound_level_pref),
    SD =sd(sound_level_pref),
    Median = median(sound_level_pref),
    SEM_max = mean(sound_level_pref) + std.error(sound_level_pref)*1.96,
    SEM_min = mean(sound_level_pref) - std.error(sound_level_pref)*1.96,
    Count = n()
  )
summary_soundlvl_bygender
```

Explain your results in plain terms here: 

Our data tells us, that the males in our sample prefer a louder sound level (a sound level of 25.6 on a scale of 0 to 100) than females in our sample (a sound level of 24.3 on a scale of 0 to 100) on average. 

To know how certain we are of the two averages we've found, we look at the 95% confidence or the interval in which we are 95% certain that the true average of the population is. In the table I have provided the maximum (SEM_max) and minimum (SEM_min) values for the 95% confidence interval for both males and females. We see that the 95% confidence interval of males incapsulate the 95% confidence interval for females. This indicates, that the difference between the two averages is not statistically significant - that the difference is likely to just have occured by chance.

We see that the medians (the value that splits the data in two equal halfs) for both males and female is 20. These values is not biased by skew and outliers as the means are, and suggests an even more equal preference of sound level. 

The two bars in the graph indicate the means of the two groups and the 'capital I's' on top of the bars indicate the 95% confidence intervals. Here we see the difference in the two means and the 95% confidence intervals of female that incapsulates the 95% confidence interval for males.

Thus, we cannot conclude from the data in our sample whether males or females prefer a louder sound level. Additionally, our data only look at 1st semester Cognitive Science students it AU, Which might or might not be representative for the entire population of males and female.

---

#### Question 3

Are the breath_hold data normally distributed? Provide visual and numeric support for your answer.

```{r question3}
#Density plot - breath hold data

ggplot(df, aes (x = breath_hold)) +
  geom_histogram(
    aes(y = ..density..),
    color = "darkblue",
    fill = "lightblue",
    binwidth = 3
  ) +
  stat_function(
    fun = dnorm,
    args = list(
      mean = mean(df$breath_hold, na.rm = TRUE),
      sd = sd(df$breath_hold, na.rm = TRUE)
    ),
    colour = "black",
    size = 1
  ) +
  labs(title = "Density plot of breath_hold data", 
       y = "Density", 
       x = "Breath hold (sec)"
  ) + 
  theme_minimal()

#QQ plot of breath hold data
  
ggplot(df, aes(sample = breath_hold)) + 
  stat_qq() + 
  stat_qq_line() +
  labs(title = "QQ plot of breath hold data", 
       y = "Breath hold (sec)", 
       x = "Expected value"
  ) + 
  theme_minimal()

#Numerical parameters - breath_hold data

round(stat.desc(df$breath_hold, basic = F, norm = T), digits = 2)
```

Explain your results in plain terms here: 

To access if the breath_hold data is normally distributed, I have provided a density plot of the breath_hold with a normal curve laid on top of it, a QQ plot of the breath_hold data, and a table with with a variety of numerical parameters.

Numerical Parameters:
For a data set to be normally distributed a couple of numerical parameters have to aply to the set. The ones I will focus on here are: the mean (average) and the median (the value that splits the data in two equal halfs) have to be (approximately) equal, the skewness (if the data is 'leaning' to one side or the other compared to a normal distribution) and kurtosis (if the tails of the data have more or fewer observations than what is to be expected in a normal distribution) devided by two times the standard error (skew.2SE and kurt.2SE) has to be within the interval from -1 to 1 preferably as close to zero as posible, and the p-value of the Shapiro-Wilk test (normtest.p) has to be greater than 0.05. The skew.2SE, the kurt.2SE, and the normtest.p values is to be interpreted causiously when the number of observations in a data set is above 30.

In the breath_hold data we have found the median to be 51.91 seconds and the mean to be 55.77 seconds. This difference implies a positive skew in the data and/or posibly one or more outliers larger that the median. But diviance from the theoretical values are to be expected and the difference is not too big for it determine if the data is normally distributed or not.
The skew.2SE has a value of 0.95, which is indicate that the data is normally distributed. The kurt.2SE has a value of -0.30, which is indicate that the data is normally distributed. The normtest.p has a value of 0.04, which is just below the acceptable level. This indicates that the data is not normally distributed. 
Because our data set has more than 30 observations (62 in fact) we cannot conclude anything definitively from the skew.2SE, the kurt.2SE, and the normtest.p values, but together with the mean-median difference these numbers suggest that the data is approximately normally distributed with some positive skew and/or outliers.

QQ-plot:
The quantile-quantile plot, or the QQ-plot, compares data colected in an experiment with the theoretically expected distribution of that data if it was normally distributed. If a data set is normally distributed the QQ-plot wil resemple a diagonal line. 

On the QQ-plot of the breath_hold data we see that the data follows the diagonal line indicating the theoretical values for a normally distributed data set. The QQ-plot of the data suggests that the breath_hold data is normally distributed.

Density plot:
On the density plot we see that the breath_hold data is slightly positively skewed, this will always be the case for time-data, because we cannot find any negative data - so zero will be a fixed lower boundry, while there is no fixed upper boundry. 

However, the black line on the density plot shows what a normally distributed data set with the same mean and standard diviation would look like. We see that the breath_hold data roughly follow this line. This suggests that the breath_hold data is normally distributed.

To sum up the assesments of normallity above: the data is approximately normal.

---

#### Question 4

Are the Balloon reaction time data (balloon and balloon_balance) normally distributed? Provide visual and numeric support for your answer.

```{r question4}
#Density plot - balloon data

ggplot(df, aes(x = balloon)) +
  geom_vline(data=df, 
             aes(xintercept = mean(balloon)+3*sd(balloon)), 
             colour="black", 
             linetype="dotted") +
  geom_histogram(
    aes(y = ..density..),
    color = "darkblue",
    fill = "lightblue",
    binwidth = 3
  ) +
  stat_function(
    fun = dnorm,
    args = list(
      mean = mean(df$balloon, na.rm = TRUE),
      sd = sd(df$balloon, na.rm = TRUE)
    ),
    colour = "black",
    size = 1
  ) + 
  labs(title = "Density Plot of Balloon Inflation Times with Normal Line", 
       y = "Density", 
       x = "Balloon Inflation Time (sec)"
       ) + 
  theme_minimal()

#Density plot - balloon_ballance data

ggplot(df, aes(x = balloon_balance)) +
  geom_vline(data=df, 
             aes(xintercept = mean(balloon)+3*sd(balloon)), 
             colour="black", 
             linetype="dotted") +
  geom_histogram(
    aes(y = ..density..),
    color = "darkblue",
    fill = "lightblue",
    binwidth = 3
  ) +
  stat_function(
    fun = dnorm,
    args = list(
      mean = mean(df$balloon, na.rm = TRUE),
      sd = sd(df$balloon, na.rm = TRUE)
    ),
    colour = "black",
    size = 1
  ) + 
  labs(title = "Density Plot of Balloon Balance Times with Normal Line",
           y = "Density", 
           x = "Balloon Ballance Time (sec)"
           ) + 
  theme_minimal()

#QQ plot - balloon data

ggplot(df, aes(sample = balloon)) + 
  stat_qq() + 
  stat_qq_line() +
  labs(title = "QQ plot of Balloon Inflation Times with Normal Line",
       y = "Balloon Inflation Time (sec)", 
       x = "Expected value"
       ) +
  theme_minimal()

#QQ plot - balloon_balance data

ggplot(df, aes(sample = balloon_balance)) + 
  stat_qq() + 
  stat_qq_line() +
  labs(title = "QQ plot of Balloon Balance Times with Normal Line",
       y = "Balloon Ballance Time (sec)", 
       x = "Expected value"
       ) +
  theme_minimal()

#Numerical parameters - balloon data

round(stat.desc(df$balloon, basic = F, norm = T), digits = 2)

#Numerical parameters - balloon_balance data

round(stat.desc(df$balloon_balance, basic = F, norm = T), digits = 2)

```


Explain your results in plain terms here: 

Balloon Inflation Data:

To access if the balloon inflation data is normally distributed, I have provided a density plot of the balloon inflation data with a normal curve laid on top of it, a QQ plot of the balloon inflation data, and a table with a variety of numerical parameters.

Numerical Parameters:
In the balloon inflation data we have found the median to be 22.50 seconds and the mean to be 33.68 seconds. This difference implies a positive skew in the data and/or posibly one or more outliers larger that the median. Diviance from the theoretical values are to be expected, but the difference is fairly large indicating that the data is not normally distributed.
The skew.2SE has a value of 3.31, which is indicate that the data has a heavy positive skew. The kurt.2SE has a value of -2.38, which is indicate that the data is not normally distributed. The normtest.p has a value of 0.00, which is below the acceptable level for the data to be normally distributed. This indicates that the data is not normally distributed. 
Because our data set has more than 30 observations (62 in fact) we cannot conclude anything definitively from the skew.2SE, the kurt.2SE, and the normtest.p values, but together with the mean-median difference these numbers suggest that the data is not normally distributed and with positive skew and/or outliers.

QQ-plot:
On the QQ-plot of the balloon inflation data we see that the data follows the diagonal line indicating the theoretical values for a normally distributed data set until just below an expected value of 1 and an inflation time of 40. From there on out the balloon inflation data deviates strongly from the expected values. The QQ-plot of the data suggests that the balloon inflation data is not normally distributed.

Density plot:
On the density plot we see a large portion of the data located below 50 seconds that actually look quite normally distributed, some observations from 50 seconds to 119 seconds, and a large amount of observations at 120 seconds (indication the time cap of the task). The density plot includes a vertical dotted line. All observations that fall above this line are outliers (the mean plus three standard diviations). There are no outliers in this data set. The black curve on the density plot shows what a normally distributed data set with the same mean and standard diviation would look like. We see that the balloon inflation data do not follow this line. 
This suggests that the balloon inflation data is not normally distributed.

To sum up the assesments of normallity above: the data is not normal.


Balloon Balance Data:

To access if the balloon balance data is normally distributed, I have provided a density plot of the balloon inflation data with a normal curve laid on top of it, a QQ plot of the balloon balance data, and a table with a variety of numerical parameters.

Numerical Parameters:
In the balloon balance data we have found the median to be 9.00 seconds and the mean to be 33.49 seconds. This difference implies a positive skew in the data and/or posibly one or more outliers larger that the median. Diviance from the theoretical values are to be expected, but the difference is very large indicating that the data is not normally distributed.
The skew.2SE has a value of 2.02, which is indicate that the data has a heavy positive skew. The kurt.2SE has a value of -0.25, which is indicate that the data has a close to normal kurtosis and is normally distributed. The normtest.p has a value of 0.00, which is below the acceptable level for the data to be normally distributed. This indicates that the data is not normally distributed. 
Because our data set has more than 30 observations (62 in fact) we cannot conclude anything definitively from the skew.2SE, the kurt.2SE, and the normtest.p values, but together with the mean-median difference these numbers suggest that the data is not normally distributed and with positive skew and/or outliers.

QQ-plot:
On the QQ-plot of the balloon balance data we see that the data does not follow the diagonal line at all indicating that the data is not normally distributed. The QQ-plot of the data suggests that the balloon balance data is not normally distributed.

Density plot:
On the density plot we see many observations close to 0 seconds, few observations from 25 seconds to 119 seconds, and a spike of observations at 120 seconds (the time cap of the task). The density plot includes a vertical dotted line. All observations that fall above this line are outliers (the mean plus three standard diviations). There are no outliers in this data set. The black curve on the density plot shows what a normally distributed data set with the same mean and standard diviation would look like. We see that the balloon balance data do not follow this line. 
This suggests that the balloon balance data is not normally distributed.

To sum up the assesments of normallity above: the data is not normal.

---

#### Question 5

Shoe size could tell us something about general body size, which could also be connected to one's ability to hold your breath. In other words we predict that there is a positive relation between shoe size and how long time CogSci students can hold their breath. Try plotting the two sets of data against each other using a scatter plot (hint: both variables are continuous variables). Bonus question: Males and females have different body sizes. Use grouping in your plot to distinguish the relationship between shoe size and holding breath for males and females.

```{r question5}
#Scatter plot - shoesize by breath_hold

ggplot(df, aes(x = shoesize, y = breath_hold)) +
  geom_point(size = 2) +
  geom_smooth(method = lm) +
  labs(title = "Scatter Plot - shoesize by breath_hold",
       y = "Breath Hold Times (sec)", 
       x = "Shoesize"
       ) + 
  theme_minimal()

#Bonus question

#Grouping the data by gender

groupedby_gender <- group_by(df, gender)

#Scatter plot - shoesize by breath_hold grouped by gender

ggplot(groupedby_gender,
       aes(x = shoesize, y = breath_hold, shape = gender)) +
  geom_point(size = 2) +
  geom_smooth(method = lm) +
  labs(title = "Scatter Plot - shoesize by breath_hold Seperated by gender", 
       y = "Breath Hold Times (sec)", 
       x = "Shoesize"
       ) +
  theme_minimal()


```

Explain your results in plain terms here: 

The Scatter Plot - shoesize by breath_hold shows how long partisipants were able to hold their breath on the x-axis and what shoesize they use on the y-axis. The trent in the scatter plot seems to be, that bigger shoesize corelates with being able to hold your breath longer, though many of the data points lays far away from the line, for example the person with the largest shoesize was one of the poorest at holding their breath. This means that the line is not very explanatory of the data.

Bonus question:

The Scatter plot - shoesize by breath_hold Seperated by gender shows how long partisipants were able to hold their breath on the x-axis and what shoesize they use on the y-axis, but this time the data has been separated into males and females. In these two groups we see two different trends. For the females it seems that there is a corelation between larger shoesize and being able to hold ones breath longer. For the males the breath hold times seem to not vary when the shoesize change. 
